Gravitational-wave sensitivity curves

### 2014-07-04 Detectors.tex
Detectors to PTAs

\subsection{Operating principle of an interferometric detector}\label{sec:principles}
All of the man-made detectors discussed in this section work on the principle of interferometry. Such detectors work by taking a beam of monochromatic light and splitting it into two beams travelling at some angle to each other. Each beam is passed in to a Fabry-P\'{e}rot cavity where it undergoes a number of round trips before being recombined to form an interference pattern. The ends of the cavity are, in the ideal case, freely floating test masses which move in response to a passing GW, this effect is measured by observing the changing interference pattern.

The response of a detector to an incident plane fronted GW depends upon the relative orientations of the detector and the incoming wave. Choose the origin of our coordinate system to be the beam-splitter of the interferometer and let $l_{1}^{i}$ and $l_{2}^{i}$ be unit 3-vectors pointing along the two arms. In the absence of noise the output of the detector is the difference in strain between the two arms \citep{MTW}
\begin{equation}\label{eq:hoftxx} h(t)=\frac{1}{2}h_{ij}\left( l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j} \right)\; , \end{equation}
where $h_{ij}$ are the spatial components of the GW metric perturbation. Let $n^{i}$ be the unit 3-vector pointing towards the source of the GWs, with spherical polar angles $(\theta,\phi)$ relative to some axes fixed to the detector, also let $p^{i}$ and $q^{i}$ be unit vectors orthogonal to $n^{i}$. There remains a freedom in the coordinates described, a rotation of $p^{i}$ and $q^{i}$ through an angle $\psi$ about $n^{i}$. Define the basis tensors,
\begin{eqnarray}
H^{+}_{ij}&=p_{i}p_{j}-q_{i}q_{j} \nonumber \\
H^{\times}_{ij}&=p_{i}q_{j}+q_{i}p_{j} \; .
\end{eqnarray}
Consider a single frequency component of the incident GW, the strain induced by the incident GW may be written as
\begin{equation}\label{eq:hijxx} h_{ij}=A_{+}H^{+}_{ij}\cos\left(2\pi ft\right)+A_{\times}H^{\times}_{ij}\cos\left(2\pi ft+\Delta \phi\right) \; ,\end{equation}
where $A_{+}$ and $A_{\times}$ are the amplitudes of the two polarisation states. Combining (\ref{eq:hoftxx}) and (\ref{eq:hijxx}) allows the detector output to be written as
\begin{equation} h(t)=F^{+}(\theta,\phi,\psi)A_{+}\cos\left(2\pi ft\right)+F^{\times}(\theta,\phi,\psi)A_{\times}\cos\left(2\pi f t + \Delta\phi \right)\; , \end{equation}
where the response functions inherit their angular dependence from the choice of coordinates,
\begin{eqnarray}\label{eq:responsefuncs}
F^{+}(\theta,\phi,\psi)&=\frac{1}{2}H^{+}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \nonumber \\
F^{\times}(\theta,\phi,\psi)&=\frac{1}{2}H^{\times}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \; . 
\end{eqnarray}
The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.

\begin{figure}[h!]
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.99\textwidth]{LIGO_detectorframe.pdf}
 \caption{The angular response function of an interferometric detector shown both as a surface plot and in an Aitoff-Hammer projection. The two detector arms lie in the $xy$ plane either side of one of the zeros in the response.}
 \label{fig:LIGO}
\end{figure}

Throughout this paper detector sensitivity refers to the polarisation and sky averaged sensitivity, $F$, where
\begin{equation}\label{eq:skyav} F^{2}=\int_{0}^{2\pi}\frac{\textrm{d}\psi}{2\pi}\; \int_{0}^{\pi}\int_{0}^{2\pi}\frac{\sin\theta\,\textrm{d}\theta\textrm{d}\phi}{4\pi}\;\left(\frac{F^{+}\left(\theta,\phi,\psi\right)^{2}+F^{\times}\left(\theta,\phi,\psi\right)^{2}}{2}\right)\; .\end{equation}
For a single $90\degree$ interferometer, such as LIGO (Laser Interferometer GW Observatory), the sky and polarisation averaged response is $F=\sqrt{1/5}\approx 0.447$.

A detector may consist of several interferometers. Let $F_{a}$ be the averaged response of the $a^{\textrm{th}}$ interferometer. The average response of the detector network is obtained by adding in quadrature,
\begin{equation} F_{\textrm{Total}}^{2}=\frac{1}{\left(\textrm{Number of interferometers}\right)}\sum_{x=1}^{N}F_{x}^{2} \; .\end{equation}

The averaging in (\ref{eq:skyav}) assumes a uniform distribution of polarisation angles $\psi$. This will be the case for a stochastic background, however for a non-inspiralling circular binary the polarisation will be a function of the two spherical polar angles ($\iota,\xi$) specifying the orientation of the binary's orbital angular momentum. Here $\iota$ is the polar angle between the orbital angular momentum and the line joining the source to the detector while $\xi$ is the azimuthal angle around the same line. In this case the correct average to characterise the detector sensitivity would be an an average over all four angles ($\theta,\phi,\iota,\xi$). If the binary is inspiralling then the polarisation will depend on still more parameters which need to be averaged over. These more complicated averages all have the property that they depend on both the detector and the source, hence they are unhelpful for our present purpose. In addition the different averages don't work out to be so different from each other; \cite{1993PhRvD..47.2198F} calculated the sensitivity for a detector with the LIGO geometry averaged over the four angles ($\theta,\phi,\iota,\xi$) as $\sqrt{4/25}=0.4$ times peak sensitivity, which should be compared with the value $\sqrt{1/5}\approx 0.447$ above. For the remainder of this paper the 3 angle average defined in (\ref{eq:skyav}) will be used.

\subsection{Ground based detectors}\label{sec:ground}
On Earth approximations to the free floating test masses described above are achieved by suspending the mass from a pendulum system with natural frequency much greater than that of the GW. The interferometric detectors listed in Table \ref{table:t} are all sensitive to GWs in the frequency range ${\cal{O}}(10-10^{3})\,\textrm{Hz}$. They all have multiple narrow lines in their sensitivity curves, which arise from noise sources in the instrument including resonances in the suspension system and electronic noise at multiples of $60\,\textrm{Hz}$: these have been removed in the \ref{app:a} figures for clarity. The detectors fall broadly into three categories: 1\textsuperscript{st} generation detectors, which have already operated; 2\textsuperscript{nd} generation detectors currently under construction; and 3\textsuperscript{rd} generation detectors at the planning stage.

\begin{table}[h!]
\caption{\label{table:t} For the AdV sensitivity curve an interpolation to the data published on \url{https://wwwcascina.virgo.infn.it/advirgo/} (2013) was used. For KAGRA (Kamioka Gravitational Wave Detector) an interpolation to the data for version D of the detector published on \url{http://gwcenter.icrr.u-tokyo.ac.jp/en/researcher/parameter} (2013) was used. For the remaining detectors analytic fits to the sensitivity curves due to \cite{Sathyaprakash} were used.}
\begin{indented}
\item[]\begin{tabular}{ l l l l l }
\br
{\bf Detector} & {\bf Country} & {\bf Arm length} & {\bf  Approximate date} & {\bf Generation} \\
\mr
  GEO600 $^{a}$	&	Germany 	& $600\,\textrm{m}$ 	& 2001-present 	   & 1$^{\textrm{st}}$\\
  TAMA300 $^{b}$& 	Japan		& $300\,\textrm{m}$ 	& 1995-present     & 1$^{\textrm{st}}$\\
  iLIGO	$^{c}$	&	US		& $4\,\textrm{km}$ 	& 2004-2010 	   & 1$^{\textrm{st}}$\\
  Virgo $^{d}$	& 	Italy		& $3\,\textrm{km}$ 	& 2007-2010 	   & 1$^{\textrm{st}}$\\
  aLIGO $^{e}$	&	US		& $4\,\textrm{km}$ 	& \emph{est.} 2015 & 2$^{\textrm{nd}}$\\
  KAGRA	$^{f}$	&	Japan		& $3\,\textrm{km}$ 	& \emph{est.} 2018 & 2$^{\textrm{nd}}$\\
  AdV $^{g}$	&	Italy	 	& $3\,\textrm{km}$ 	& \emph{est.} 2016 & 2$^{\textrm{nd}}$\\
  ET (Einstein Telescope)$^{h}$ & Italy		& $10\,\textrm{km}$ 	& \emph{est.} 2025 & 3$^{\textrm{rd}}$\\
\br
\multicolumn{5}{l}{$^{a}$\cite{2008CQGra..25k4043G}, $^{b}$\cite{2002CQGra..19.1409A}, }\\
\multicolumn{5}{l}{$^{c}$\cite{2009RPPh...72g6901A}, $^{d}$\cite{accadia_virgo:_2012}, $^{e}$\cite{2010CQGra..27h4006H}, }\\
\multicolumn{5}{l}{$^{f}$\cite{2012CQGra..29l4007S}, $^{g}$\cite{Acernese2009}, $^{h}$\cite{2011CQGra..28i4013H}.}\\
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Space based detectors}\label{sec:space}
Space based detectors work on similar principles to ground based detectors but with the test masses residing inside of independent, widely separated satellites. Space based detectors are sensitive to lower frequency GWs than their ground based counterparts, this is partly because space based detectors can have much longer arms and partly because they are unaffected by seismic noise which limits the low frequency performance of ground based detectors.

The canonical design for a space based detector is LISA (Laser Interferometer Space Antenna), which would be sensitive to milli-Hz GWs. LISA would consist of three satellites flying in a triangular constellation with arms of length $5\times 10^{9}\,\textrm{m}$ in a $1\,\textrm{AU}$ orbit around the sun, trailing the Earth by $20\degree$. The laser arms in a LISA like detector are not a Fabry-P\'{e}rot cavity, the light only travels once along each arm. All of the sources discussed here lie within LISA's sensitivity reach. eLISA (evolved LISA) is a simpler version of LISA designed to probe the same frequency range, while proposals such as ALIA (Advanced Laser Interferometer Antenna), BBO (Big Bang Observer) and DECIGO (DECI-Hertz Interferometer GW Observatory) are designed to probe deci-Hz GWs.

\subsubsection{LISA and eLISA}
We use an analytic fit to the instrumental noise curve given by \cite{Sathyaprakash}. When observing individual sources with LISA there is an additional contribution to the noise from a background of unresolvable binaries. This is not included here as we consider the background as a source of GWs, see section \ref{sec:GB}. eLISA is a re-scoped version of the classic LISA mission, the main differences are shorter arms ($10^{9}\,\textrm{m}$ instead of $5\times 10^{9}\,\textrm{m}$), two laser arms instead of three and a different orbit (drifting away from Earth instead of $20\degree$ Earth trailing). The effect of these changes is a slightly reduced peak sensitivity and a shift to higher frequencies. We use an analytic fit to the instrumental noise curve given by \cite{Amaro-Seoane-et-al}.

\subsubsection{ALIA, DECIGO and BBO}
These missions are designed to probe the deci-Hz region of the GW spectrum, they are considerably more ambitious than the LISA or eLISA mission and their launches will be further into the future. For ALIA a simple analytic fit to the sensitivity curve is used, while for DECIGO and BBO fits to the sensitivity curves given by \cite{2011PhRvD..83d4011Y} are used.

@@

\begin{equation}\label{eq:pulsarredshift} z(t,\hat{\Omega})=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\left(h_{ij}^{\textrm{Pulsar}}(t-L,\hat{\Omega} )-h_{ij}^{\textrm{Earth}}(t,\hat{\Omega} )\right)=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\Delta h_{ij}(t,\hat{\Omega})\; . \end{equation}

@@

\begin{equation}\label{eq:restored}  R(t,\hat{\Omega})=\int_{0}^{t}\textrm{d}t'\;z(t',\hat{\Omega}) \; . \end{equation}

@@

\begin{equation} h_{ij}(t,\hat{x})=\sum_{A=+,\times}\int\textrm{d}f\;\iint_{S_{2}}\textrm{d}\hat{\Omega}\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f (t-\hat{\Omega}\cdot\vec{x}) } \; ,\end{equation}

@@

\begin{equation}\label{eq:deltahij} \Delta h_{ij}(t,\hat{\Omega})=\sum_{A=+,\times}\int\textrm{d}f\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f t }\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right) \; . \end{equation}

@@

\begin{equation}\label{eq:45} \tilde{z}(f,\hat{\Omega})=\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right)\sum_{A=+,\times} \tilde{h}_{A}(f,\hat{\Omega})F^{A}(\hat{\Omega})\;, \textrm{where}\; F^{A}(\hat{\Omega})=\frac{e_{ij}^{A}(\hat{\Omega})\hat{p}^{j}\hat{p}^{i}} {2(1+\hat{\Omega}\cdot\hat{p})} \, .\end{equation}

@@

\begin{eqnarray} & \left<z_{1}(f)z_{2}^{*}(f')\right>=\frac{1}{2}S_{h}(f)\delta (f-f')\Gamma(f)\quad \textrm{where,} \\
& \Gamma(f)=\sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\textrm{d}\hat{\Omega}\,\left(e^{2\pi\rmi fL_{1}(1+\hat{\Omega}\cdot\hat{p}_{1})}-1\right)\left(e^{-2\pi\rmi fL_{2}(1+\hat{\Omega}\cdot\hat{p}_{2})}-1\right)F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega}) \; . \nonumber\end{eqnarray}

@@

\begin{equation} \Gamma(f)\approx\Gamma_{0}= \sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\textrm{d}\hat{\Omega}\;F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega})\;.\end{equation}

@@

\begin{equation} \Gamma_{0}=\frac{1}{2}+\frac{3x}{2}\left(\log (x)-\frac{1}{6}\right) \quad \textrm{where}\quad x=\frac{1-\cos\theta}{2}\; .\end{equation}

@@

\subsubsection{SKA}
The sensitivity curves plotted in \ref{app:a} for the SKA (Square Kilometer Array) assume values of $\delta t^{2}_{\textrm{r.m.s.}}$ a factor of 10 better than current PTAs and 20 pulsars timed for twice as long as the current EPTA.

>>>>

\subsection{Operating principle of an interferometric detector}\label{sec:principles}

All of the man-made detectors discussed in this section work on the principle of interferometry. Such detectors work by taking a beam of monochromatic light and splitting it into two beams travelling at some angle to each other. Each beam is passed in to an optical cavity where it undergoes a number of round trips before being recombined to form an interference pattern. The ends of the cavity are, in the ideal case, freely floating test masses which move in response to a passing GW, this effect is measured by observing the changing interference pattern. % LIGO isn't Fabry-P\'{e}rot

The response of a detector to an incident plane-fronted GW depends upon the relative orientations of the detector and the incoming wave. Let us choose the origin of our coordinate system to be the beam-splitter of the interferometer, and $l_{1}^{i}$ and $l_{2}^{i}$ to be unit 3-vectors pointing along the two arms. In the absence of noise the output of the detector is the difference in strain between the two arms \citep{Thorne1987} \cplb{I think you mean this rather than MTW}
\begin{equation}\label{eq:hoftxx}
h(t)=\frac{1}{2}h_{ij}\left( l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j} \right)\; ,
\end{equation}
where $h_{ij}$ are the spatial components of the GW metric perturbation. Let $\hat{r}^{i}$ be the unit 3-vector pointing towards the source of the GWs, with spherical polar angles $(\theta,\phi)$ relative to some axes fixed to the detector, and let $p^{i}$ and $q^{i}$ be unit vectors orthogonal to $\hat{r}^{i}$. We can now define the basis tensors
\begin{eqnarray}
H^{+}_{ij}&=p_{i}p_{j}-q_{i}q_{j} \nonumber \\
H^{\times}_{ij}&=p_{i}q_{j}+q_{i}p_{j} \; .
\end{eqnarray}
There remains a freedom in the coordinates described, a rotation of $p^{i}$ and $q^{i}$ through an angle $\psi$ about $\hat{r}^{i}$ known as the polarization angle. For a single frequency component, the strain induced by GW may be written as
\begin{equation}\label{eq:hijxx}
h_{ij}=A_{+}H^{+}_{ij}\cos\left(2\pi ft\right)+A_{\times}H^{\times}_{ij}\cos\left(2\pi ft+\Delta \phi\right) \; ,
\end{equation}
where $A_{+}$ and $A_{\times}$ are the amplitudes of the two polarisation states. Combining (\ref{eq:hoftxx}) and (\ref{eq:hijxx}) allows the detector output to be written as
\begin{equation}
h(t)=F^{+}(\theta,\phi,\psi)A_{+}\cos\left(2\pi ft\right)+F^{\times}(\theta,\phi,\psi)A_{\times}\cos\left(2\pi f t + \Delta\phi \right)\; ,
\end{equation}
where the response functions inherit their angular dependence from the choice of coordinates,
\begin{eqnarray}\label{eq:responsefuncs}
F^{+}(\theta,\phi,\psi)&=\frac{1}{2}H^{+}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \; , \\
F^{\times}(\theta,\phi,\psi)&=\frac{1}{2}H^{\times}_{ij}\left(l_{1}^{i}l_{1}^{j}-l_{2}^{i}l_{2}^{j}\right) \; . 
\end{eqnarray}
The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.

\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.99\textwidth]{LIGO_detectorframe.pdf}
 \caption{The angular response function of an interferometric detector shown both as a surface plot and in an Aitoff--Hammer projection \cplb{[citation needed]}. The two detector arms lie in the $x$--$y$ plane either side of one of the zeros in the response. \cplb{Can you label the colour bar/explain what the surface indicates?}}
 \label{fig:LIGO}
\end{figure}

Throughout this paper detector sensitivity refers to the polarisation and sky averaged sensitivity $F$, where
\begin{equation}\label{eq:skyav}
F^{2}=\int_{0}^{2\pi}\frac{\mathrm{d}\psi}{2\pi}\; \int_{0}^{2\pi} \frac{\mathrm{d}\phi}{2\pi}\; \int_{0}^{\pi}\frac{\sin\theta\,\mathrm{d}\theta}{2}\;\left[\frac{F^{+}\left(\theta,\phi,\psi\right)^{2}+F^{\times}\left(\theta,\phi,\psi\right)^{2}}{2}\right]\; .
\end{equation}
For a single $90\degree$-interferometer, such as LIGO, the sky and polarisation averaged response is $F=\sqrt{1/5}\approx 0.447$.

A detector may consist of several interferometers. Let $F_{a}$ be the averaged response of the $a$-th interferometer, the average response of a network of $k$ detectors is obtained by adding in quadrature,
\begin{equation} F_{\mathrm{Total}}^{2}=\frac{1}{k)}\sum_{a\,=\,1}^{k}F_{a}^{2} \; .\end{equation}

The averaging in (\ref{eq:skyav}) assumes a uniform distribution of polarisation angles $\psi$. This is the case for a stochastic background; however, for a non-inspiralling circular binary, the polarisation is a function of the two spherical polar angles ($\iota,\xi$) specifying the orientation of the binary's orbital angular momentum. Here, $\iota$ is the inclination angle, the polar angle between the orbital angular momentum and the line joining the source to the detector \cplb{($-\hat{r}$)} and $\xi$ is the azimuthal angle around the same line. In this case, to characterise the detector sensitivity we average over all four angles ($\theta,\phi,\iota,\xi$). If the binary is inspiralling, then the polarisation depends on still more parameters which need to be averaged over. These more complicated averages all have the property that they depend on both the detector and the source, hence they are unhelpful for our present purpose \cplb{of \ldots}. Additionally, the different averages do not work out to be so different from each other: \citet{1993PhRvD..47.2198F} calculated the sensitivity for a detector with the LIGO geometry averaged over the four angles ($\theta,\phi,\iota,\xi$) as $\sqrt{4/25}=0.4$ times peak sensitivity, which should be compared with the value $\sqrt{1/5}\approx 0.447$ above. For the remainder of this paper the three-angle average defined in (\ref{eq:skyav}) is used.

\subsection{Ground-based detectors}\label{sec:ground}

Ground-based detectors are the most numerous. A collection of interferometric detectors are listed in table \ref{table:t}, these are sensitive to GWs in the frequency range ${\mathcal{O}}(10$--$10^{3})~\mathrm{Hz}$. They all simulate free-floating test masses by suspending a mass from a pendulum system with natural frequency much greater than that of the GW. Their sensitivity curves include narrow lines, which arise from noise sources in the instrument including resonances in the suspension system and electrical noise at multiples of $60~\mathrm{Hz}$: these have been removed in the \ref{app:a} figures for clarity. The detectors fall broadly into three categories: first-generation detectors, which have already operated; second-generation detectors currently under construction, and third-generation detectors at the planning stage.

The most notable ground-based detectors are LIGO and Virgo \citep{Abramovici1992,Acernese2008a}. These work in collaboration, supported by GEO600 \citep{Willke2002}. LIGO and Virgo have completed science runs as first-generation detectors, now referred to as Initial LIGO (iLIGO) and Initial Virgo (iVirgo). They are now being upgraded to second-generation detectors, advanced LIGO (aLIGO) and advanced Virgo (AdV). \cplb{I think AdV is the official acronym, although it does break the pattern.} LIGO has two observatories, one at Hanford, Washington, which has two detectors, and another at Livingston, Louisiana. There is an agreement to move one of the upgraded Hanford detector systems to a location in India \citep{LIGO-India,Unnikrishnan2013}. The advanced detectors should start operation in the next couple of years, with LIGO-India following further in the future.

TAMA300 is a Japanese first-generation detector. Its successor is the Kamioka Gravitational Wave Detector (KAGRA), formerly the Large-scale Cryogenic Gravitational wave Telescope (LCGT), which is located underground in the Kamioka mine. It employs more sophisticated noise-reduction techniques than LIGO or Virgo, such as cryogenic cooling \citep{2012CQGra..29l4007S}.

The Einstein Telescope (ET) is an ambitious proposal to construct an underground third-generation detector \citep{Punturo2010,Sathyaprakash2012}. Its location would provide shielding from seismic noise, allowing it to observe frequencies of $10$--$10^4~\mathrm{Hz}$.

\begin{table}
\caption{\label{table:t} Summary of ground-based laser interferometers.}
\begin{indented}
\item[]\begin{tabular}{ l l l l l }
\br
{\bf Detector} & {\bf Country} & {\bf Arm length} & {\bf  Approximate date} & {\bf Generation} \\
\mr
  GEO600$^{a}$	&	Germany 	& $600~\mathrm{m}$ 	& 2001--present	   & First \\
  TAMA300$^{b}$ & 	Japan		& $300~\mathrm{m}$ 	& 1995--present    & First \\
  iLIGO$^{c}$	&	USA		& $4~\mathrm{km}$ 	& 2004--2010 	   & First \\
  Virgo$^{d}$	& 	Italy		& $3~\mathrm{km}$ 	& 2007--2010 	   & First \\
  aLIGO$^{e}$	&	USA		& $4~\mathrm{km}$ 	& \emph{est.} 2015 & Second \\
  AdV$^{f}$	&	Italy	 	& $3~\mathrm{km}$ 	& \emph{est.} 2016 & Second \\
  KAGRA$^{g}$	&	Japan		& $3~\mathrm{km}$ 	& \emph{est.} 2018 & Second \\
  ET$^{h}$      & 	---		& $10~\mathrm{km}$ 	& \emph{est.} 2025 & Third \\
\br
\multicolumn{5}{l}{$^{a}$\citet{2008CQGra..25k4043G}, $^{b}$\citet{2002CQGra..19.1409A}, $^{c}$\citet{2009RPPh...72g6901A}, $^{d}$\citet{accadia_virgo:2012}, }\\
\multicolumn{5}{l}{$^{e}$\citet{2010CQGra..27h4006H}, $^{f}$\citet{Acernese2009}, $^{g}$\citet{2012CQGra..29l4007S}, $^{h}$\citet{2011CQGra..28i4013H}.}\\
\br
\end{tabular}
\end{indented}
\end{table}

In \ref{app:a}, we use an interpolation to the data published on \url{https://wwwcascina.virgo.infn.it/advirgo/} (2013) for the AdV sensitivity curve, an interpolation to the data for version D of the detector published on \url{http://gwcenter.icrr.u-tokyo.ac.jp/en/researcher/parameter} (2013) for KAGRA and analytic fits to the sensitivity curves from \citet{Sathyaprakash} for the remaining detectors.

\subsection{Space-based detectors}\label{sec:space}

Space-based detectors work on similar principles to ground based detectors, but with the test masses residing inside of independent, widely separated satellites. Space-based detectors are sensitive to lower frequency GWs than their ground-based counterparts; this is partly because space based detectors can have much longer arms, and partly because they are unaffected by seismic noise which limits the low frequency performance of ground-based detectors.

The canonical design for a space-based detector is the aser Interferometer Space Antenna (LISA), which is sensitive to millihertz GWs. LISA would consist of three satellites flying in a triangular constellation with arms of length $5\times 10^{9}~\mathrm{m}$ in a $1~\mathrm{AU}$ orbit around the sun, trailing the Earth by $20\degree$. The laser arms in a LISA like detector are not a cavity, the light only travels once along each arm. eLISA is a simpler version of LISA designed to probe the same frequency range, while proposals such as the Advanced Laser Interferometer Antenna (ALIA), Big Bang Observer (BBO) and Deci-hertz Interferometer GW Observatory (DECIGO) are designed to probe decihertz GWs.

\subsubsection{LISA and eLISA}

The instrumental noise curves for LISA by the analytic fit given by \citet{Sathyaprakash}, which we use for the plots in \ref{app:a}. When observing individual sources with LISA there is an additional contribution to the noise from a background of unresolvable binaries. This is not included here as we consider the background as a source of GWs (see section \ref{sec:GB}). eLISA is a re-scoped version of the classic LISA mission, the main differences are shorter arms ($10^{9}~\mathrm{m}$ instead of $5\times 10^{9}~\mathrm{m}$), two laser arms instead of three, and a different orbit (drifting away from Earth instead of $20\degree$ Earth trailing). The effect of these changes is a slightly reduced peak sensitivity and a shift to higher frequencies. In \ref{app:a}, we use an analytic fit to the instrumental noise curve given by \citet{Amaro-Seoane-et-al}.

\subsubsection{DECIGO, ALIA and BBO}

These missions are designed to probe the decihertz region of the GW spectrum; they are considerably more ambitious than the LISA or eLISA mission and their launches will be further into the future. In \ref{app:a} we use a simple analytic fit to the sensitivity curve for ALIA \cplb{[citation needed]}, while for DECIGO and BBO, fits to the sensitivity curves given by \citet{2011PhRvD..83d4011Y} are used.

@@

\begin{equation}\label{eq:pulsarredshift} z(t,\hat{\Omega})=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\left(h_{ij}^{\mathrm{Pulsar}}(t-L,\hat{\Omega} )-h_{ij}^{\mathrm{Earth}}(t,\hat{\Omega} )\right)=\frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\Delta h_{ij}(t,\hat{\Omega})\; . \end{equation}

@@

\begin{equation}\label{eq:restored}  R(t,\hat{\Omega})=\int_{0}^{t}\mathrm{d}t'\;z(t',\hat{\Omega}) \; . \end{equation}

@@

\begin{equation} h_{ij}(t,\hat{x})=\sum_{A=+,\times}\int\mathrm{d}f\;\iint_{S_{2}}\mathrm{d}\hat{\Omega}\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f (t-\hat{\Omega}\cdot\vec{x}) } \; ,\end{equation}

@@

\begin{equation}\label{eq:deltahij} \Delta h_{ij}(t,\hat{\Omega})=\sum_{A=+,\times}\int\mathrm{d}f\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega})e^{ 2\pi i f t }\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right) \; . \end{equation}

@@

\begin{equation}\label{eq:45} \tilde{z}(f,\hat{\Omega})=\left(e^{-2\pi\rmi fL (1+\hat{p}\cdot\hat{\Omega})}-1\right)\sum_{A=+,\times} \tilde{h}_{A}(f,\hat{\Omega})F^{A}(\hat{\Omega})\;, \mathrm{where}\; F^{A}(\hat{\Omega})=\frac{e_{ij}^{A}(\hat{\Omega})\hat{p}^{j}\hat{p}^{i}} {2(1+\hat{\Omega}\cdot\hat{p})} \, .\end{equation}

@@

\begin{eqnarray} & \left<z_{1}(f)z_{2}^{*}(f')\right>=\frac{1}{2}S_{h}(f)\delta (f-f')\Gamma(f)\quad \mathrm{where,} \\
& \Gamma(f)=\sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\mathrm{d}\hat{\Omega}\,\left(e^{2\pi\rmi fL_{1}(1+\hat{\Omega}\cdot\hat{p}_{1})}-1\right)\left(e^{-2\pi\rmi fL_{2}(1+\hat{\Omega}\cdot\hat{p}_{2})}-1\right)F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega}) \; . \nonumber\end{eqnarray}

@@

\begin{equation} \Gamma(f)\approx\Gamma_{0}= \sum_{A=+,\times}\iint_{{\cal{S}}^{2}}\mathrm{d}\hat{\Omega}\;F_{1}^{A}(\hat{\Omega})F_{2}^{A}(\hat{\Omega})\;.\end{equation}

@@

\begin{equation} \Gamma_{0}=\frac{1}{2}+\frac{3x}{2}\left(\log (x)-\frac{1}{6}\right) \quad \mathrm{where}\quad x=\frac{1-\cos\theta}{2}\; .\end{equation}

@@

\subsubsection{SKA}
The sensitivity curves plotted in \ref{app:a} for the SKA (Square Kilometer Array) assume values of $\delta t^{2}_{\mathrm{r.m.s.}}$ a factor of 10 better than current PTAs and 20 pulsars timed for twice as long as the current EPTA.
